---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Yuqi Zhang"
date: 2026-09-08
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, In this assignment, I examine median gross rent across
counties and county-equivalents in Virginia. I compare their margins of error to
identify which communities are measured more reliably and which have greater
uncertainty.

---

## Choosing what to measure

```{r}
my_state <- "VA"        # two-letter abbreviation, in quotes
my_variable <- "B25064_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

**The dataset contains 133 rows, representing Virginia’s 95 counties and 38 independent cities, which are treated as county-equivalent geographies in the ACS.**

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```


Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

**Surry County has the largest margin of error relative to its estimate. Its margin of error is approximately 30.0% of the estimated median gross rent, making it the least reliably measured geography in this dataset.**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

**There are 49 county-equivalent geographies in the high-confidence category, 57 in the moderate-confidence category, and 27 in the low-confidence category.**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row. This requires a wide-format ACS pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```

```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE, na.rm = TRUE)
  )
```

**High-confidence counties have an average population of approximately 134,521, while low-confidence counties have an average population of only 12,508. Moderate-confidence counties fall in between, with an average population of about 26,808. This shows a clear pattern: counties with smaller populations tend to have larger margins of error relative to their estimates.**

---

## What this means for policy

**The ACS data show that we know the least about some smaller communities in Virginia when measuring median gross rent. Surry County has the largest relative margin of error at 30.0%, followed by Greene County at 27.2%, Covington city at 26.8%, and Powhatan County at 26.5%. These large margins of error mean that their rent estimates are less precise than estimates for communities with smaller margins of error.

This pattern is related to how the ACS works. The ACS is based on a sample rather than a complete count of every household. Smaller communities generally have smaller samples, which creates more uncertainty in their estimates. My results show this pattern clearly. High-confidence counties have an average population of about 134,521, while low-confidence counties have an average population of only 12,508. Moderate-confidence counties fall in between, with an average population of about 26,808.

This uncertainty could cause problems if a state agency used median gross rent to allocate affordable housing funding or rental assistance while treating every estimate as equally certain. Smaller communities with large margins of error would be especially vulnerable. For example, if their estimated rent is lower than the actual rent, they might receive less housing assistance than they actually need. If it is higher than the actual rent, funding might be directed there instead of to communities with greater needs. Therefore, agencies should consider both ACS estimates and their margins of error when making funding decisions.**



